---
title: "lab_04"
author: "Derek Willis"
date: "2023-10-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

You'll need to load three packages for this: the tidyverse, lubridate and janitor.

**Task** load these three packages.

```{r}
# Turn off scientific notation
options(scipen=999)

# load your packages
library(tidyverse)
library(lubridate)
library(janitor)
```

For this lab, we want to investigate spending by Maryland state and local candidates during the previous election. For example, we could ask how much money have candidates and committees spent on advertising on Facebook? Which vendors received the most money for media expenses? We have the data, but it's messy - names of payees are spelled differently and capitalized differently - and we need to clean it up. We'll use a combination of RStudio and OpenRefine to do that.

The basic process here is to start in RStudio, export data so that we can use OpenRefine to clean it up, and then bring it back into RStudio to finish our analysis.

## Load Data

You'll need to load one data set, a CSV file of Maryland campaign expenditures located in the data folder called "maryland_expenses.csv"

**Task** Create a codeblock below, then read the data in and assign it to an appropriate variable name. You'll want to clean up the column names and make sure that any date columns are actually date datatypes.

```{r}
# Load the Maryland expenditures data table
md_campaign_exp <- read_csv("data/maryland_expenses.csv")

clean_md_campaign_exp <- md_campaign_exp |> clean_names() |> mutate(expenditure_date=mdy(expenditure_date))

clean_md_campaign_exp
```

## Answer questions

**Q1.** You've been assigned to write about Maryland political committees' use of Facebook advertising. There's a category column in the data that has values for different kinds of spending. One of them is "Media". You'll need to complete the following steps:

   1. Create a dataframe of Media expenses where the name of the payee is not NA.
   2. Then, using `write_csv`, make a CSV file of your media expenses in your data folder. `write_csv()` is similar to `read_csv()` but you provide it with the dataframe you're writing and the file path.
   3. Using OpenRefine, create a project using that CSV file, create a copy of the `payee_name` column called `payee_clean` and then standardize it, focusing on the payees that appear most often. Pay particular attention to variations of Facebook, bearing in mind that it changed its name in 2021: <https://about.fb.com/news/2021/10/facebook-company-is-now-meta/>), but also clean up variations of the most common payees.
   4. When you are done standardizing `payee_clean`, export it as a new CSV file to your data folder, *giving it a different name* and read it back into RStudio as a new dataframe.
   5. Using that `payee_clean` column in your new dataframe, calculate the total amount of Facebook spending by each committee and present the results showing the committee that spent the most first.

**A1.**

```{r}
media_exp <- clean_md_campaign_exp |> filter(expense_category=="Media" & payee_name!="NA")

write_csv(media_exp, "data/media_exp.csv")

clean_media_exp <- read_csv("data/clean-media-exp.csv")

facebook_spending <- clean_media_exp |> group_by(committee_name) |> filter(payee_clean=="Facebook/Meta") |> summarize(total_amount = sum(amount)) |> arrange(desc(total_amount))

facebook_spending
```

**Q2.** Let's do some research on the top committees from question 1. Look up the top 10 spenders: if they are candidate committees, what offices were they running for? If they were not candidates, what issues do they seem to care about? Write some sentences answering those questions, then go the Facebook ad library ([https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\\[direction\\]=desc&sort_data\\[mode\\]=relevancy_monthly_grouped&media_type=all](https://www.facebook.com/ads/library/?active_status=all&ad_type=political_and_issue_ads&country=US&sort_data\%5Bdirection\%5D=desc&sort_data\%5Bmode\%5D=relevancy_monthly_grouped&media_type=all){.uri}) and look for ads from the top spender. What do you notice about the the information provided about those ads?

**A2.** 
**The top spenders were the friends of Ashwani Jain - a UMD alumnus running for Congress for the 6th district of Maryland. The second top spender was Peter V.R. Franchot  & Monique Anderson-Walker's joint campaign for Governor and Lt. Governor, closely followed by the friends of Franchot. The fourth leading spender of Facebook ads was the Casa in Action PAC - a 501(c3) that focuses on mobilizing Black and brown voters and "electing progressive change leaders". In fifth place, the friends of Billy Boniface - the Director of Administration for Harford County Government for the past 8 years. The current top spenders of Facebook ads are the Electronic Payments Coalition, American Petroleum Institute, Americans for Prosperity Action, WorkMoney, and Biden for President.**

**Q3.** Given the list of top spenders in A1 and the primary results for governor (<https://elections.maryland.gov/elections/2022/primary_results/gen_results_2022_1.html>), who is missing from that Q1 list? Did those campaigns run Facebook ads? Who are the biggest recipients of media expenses by the Democratic primary winner for governor (HINT: that person has two committees, the second one with his running mate)?

**A3.**
**Wes Moore and Aruna Miller is missing from the results of Q1. They did not run any Facebook ads however there were ads ran on the platform by Casa in Action PAC in support of Miller. Moore's biggest media expense was to SKDK with a total of $6,172,801.71 spent.**
```{r}
clean_media_exp |>
  filter(committee_name=="Moore  Wes For Maryland" | committee_name=="Moore Miller For Maryland Slate") |>
  group_by(payee_name) |>
  summarize(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```

**Q4.** The media expense data you cleaned also has a standardized purpose column. Write code to generate a list of the committees with the total amount of spending on online advertising. How does this list compare to the one from Q1 about top Facebook spenders? Of the top 5 committees who spent the most on Facebook ads from Q1, are there any that appear to have spent *all* of their online advertising dollars on Facebook?

**A4.** 
**Casa in Action PAC is still one of the top five spenders in comparison to the Q1 results. Of the top 5 Facebook spenders, both the friends of Ashwani Jain and the joint campaign for Franchot and Anderson-Walker spent 100% of their online advertising funds on Facebook. The friends of Peter Franchot spent 80% of their funds on Facebook ads. The friends of Billy Boniface spent 65% while Casa in Action PAC spent nearly 4% of their funds on the platform.**
```{r}
online_ads <- clean_media_exp |>
 group_by(committee_name) |>
 filter(expense_purpose=="Online Advertising") |>
 summarize(total_amount = sum(amount)) |>
 arrange(desc(total_amount))

online_ads
```

**Q5** Your editor asks you to summarize your findings about spending on Facebook by Maryland political candidates. How much do we know and what don't we know? What's the best story idea you've found here?

**A5**
**I think it's worth noting that those that won the primary election, Republican and Democrat, did not put ANY of their online advertising funds towards Facebook advertising. It could definitely call to question the effectiveness of Facebook as an advertising platform in the state and potentially shine light on other more effective platforms that Marylanders are looking at.**