---
title: "lab_02"
author: "derek willis"
date: "2023-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with the UMD courses and Maryland campaign expense data we used in the pre_lab.

```{r}
umd_courses <- read_rds("data/umd_courses.rds")
maryland_expenses <- read_csv("data/maryland_expenses.csv")
```

## Answer questions

**Q1.** How many departments have offered the courses with "Climate" in the title and at least one seat? Which department has offered the most? **A1.**
**A total of 26 departments offered courses with at least one seat and "Climate" in the title. The Atmospheric and Oceanic Science department offered the most of these courses with 68 total course offerings.**
```{r}
umd_courses |>
  filter(str_detect(title, "Climate")) |>
  filter(seats >= 1) |>
  group_by(department) |>
  summarize(count_classes = n()) |>
  arrange(desc(count_classes))
```

**Q2.** Of the departments listed in A1, which one is the most interesting to you, and why? 
**A2.**
**Art History & Archaeology was the most interesting to me because I can't imagine how the courses offered could incorporate climate.**
**Q3** Which departments offered the most courses in the Fall 2023 term? And which departments in that term have at least one syllabus posted? Answer both of those questions and then compare the top five departments - how are they different?

**A3.**
**Business and Management offers the most courses with one syllabus posted with 82 course offerings. In comparison to the top five departments, Business and Management is a significant outlier as the other four average around 45 course offerings.**
```{r}
umd_courses |>
 filter(syllabus_count >= 1) |>
 filter(str_detect(term, "202308")) |>
 group_by(department) |>
 summarize(count_classes = n()) |>
 arrange(desc(count_classes))
```

**Q4.** Using the Maryland campaign expenses data, change the datatype of the expenditure_date column so that it is a date and not a chr, saving the result as a new dataframe. Then create a new column for the month based on expenditure_date. Then write code that shows total amount spent each month by Wes Moore's committee (you'll need to search the data to find the exact name, and we're not looking in payee_name). Write a sentence describing the pattern of spending over time. 
**A4.**
**Moore_expenses <- maryland_expenses |> mutate(expenditure_date = mdy(expenditure_date))**
**Expenses are fairly consistent throughout the progression of time however, in June, July, and October of 2022 spending had a significant spike.**
```{r}
Moore_expenses |>
 mutate(month = floor_date(expenditure_date, "month")) |>
 group_by(month) |>
 filter(str_detect(committee_name, "Moore  Wes For Maryland")) |>
 summarize(total_amount = sum(amount)) |>
 arrange(desc(month))
```

**Q5.** Using case_when(), create a column in the Maryland expenses data called `spending_location` indicating whether each record spent money in Maryland or outside Maryland. For records that are in Maryland, make the new column's value "In-state" and for all others, make it "Out of state". Then write code that shows the total amount of money spent in each category and describe the results. You can do this in one statement or break it up into two statements (saving the new column to your dataframe).
**A5**
**Although the raw data shows a number of out-of-state expenses, the system was unable to calculate the sum of those expenses. However, it was able to calculate the sum of the in-state expenses with a total of $77,723,146.**
```{r}
maryland_expenses_with_state <- maryland_expenses |>
    mutate(spending_location = case_when(str_detect(address, " Maryland ") ~ "in-state", .default = "out-of-state")) |>
    group_by(spending_location) |>
    summarize(total_amount = sum(amount))
```

**Q6.** Choose your own adventure: using either the UMD courses or campaign expenses data, ask and answer a question that is different than the previous ones. You can use one of the ones you provided in the pre_lab. 
**A6.**
**3.Who expensed the largest candidate loan payment in 2022?**
**Jonathan Baron paid the largest candidate loan payment in 2022 with a payment of $299,500.**
```{r}
candidate_loan_payments <- maryland_expenses |> mutate(expenditure_date = mdy(expenditure_date)) |> mutate(year = floor_date(expenditure_date, "year")) |> filter(expense_category == "Candidate Loan Payment" & year == "2022-01-01") |> select(payee_name, amount) |> arrange(desc(amount))
```
