---
title: "lab_09.Rmd"
author: "derek willis"
date: "2023-04-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* A Datawrapper account
* Data in CSV files in the `data` folder

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab.

```{r}
# load libraries here
library(tidyverse)
library(janitor)
```

## Questions

**Q1.** You want to explore the differences between the electoral performances of Dan Cox in 2022 and Larry Hogan in 2018. Except a map is probably not the best way to show that. Let's make a chart instead, and the type is up to you, but the goal is to clearly show the differences between the two.

You MUST do the following things when answering this question:
* Load the md_gov_county.csv and md_gov_county_18.csv files from the data folder
* Join those dataframes, specifying the join column. 
* Add columns to generate percentages for Cox and Hogan.
* Export that joined dataframe to a CSV file using `write_csv`.
* Using that exported CSV, make a chart with Datawrapper. What kind of chart is up to you, so experiment with a few kinds of charts to see what answers the question above in the clearest way. Your chart MUST:
  * Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.
  * Easily readable labels and values
* Publish your chart
* Describe your thought process below - why you chose to make the chart you did - and add the URL of your chart.

**A1.** 
**https://datawrapper.dwcdn.net/fkTyT/1/ - I chose a grouped bar chart because it grouped the results of the county in an easy way for the reader's to compare the result of Hogan's 2018 campaign and Cox's 2022 campaign without having to over analyze the graphic to understand the trends. It was also for this reason I chose to sort the chart by Hogan's data - to really point out how many more votes he had over Cox. I wasn't sure whether to go with the number of votes each candidate received or the percentage of all votes their campaigns received. I ultimately went with the percentages because I felt it was easier on the eye than the number of votes themselves. When analyzing the data, the only thing that really stood out to me was how much of a difference there was between Hogan and Cox's votes/percentages in every county so I chose that as my headline and further explained how stark the difference was in the caption. I felt it was important to explain that Saint Mary's County did in fact have data but, since there was no 2022 data to compare it to, there was no percentage value that could be calculated.**
```{r}
#Load the md_gov_county.csv and md_gov_county_18.csv files from the data folder
md_gov_county_22 <- read_csv("data/md_gov_county.csv")
md_gov_county_22 <- md_gov_county_22 |> 
  clean_names()
md_gov_county_22
md_gov_county_18 <- read_csv("data/md_gov_county_18.csv")
md_gov_county_18 <- md_gov_county_18 |> 
  clean_names()
md_gov_county_18
#Join those dataframes, specifying the join column
md_gov_combo <- left_join(md_gov_county_18, md_gov_county_22, by = c("geoid", "county"))
md_gov_combo
#Add columns to generate percentages for Cox and Hogan
hogan_v_cox <- md_gov_combo |> 
  mutate(
    total_votes = hogan + jealous + quinn + schlakman + write_ins.x + cox + moore + lashar + wallace + harding + sefcik + write_ins.y,
    pct_hogan = (hogan/total_votes)*100,
    pct_cox = (cox/total_votes)*100
  ) |>
  group_by(county) |>
  select(county, hogan, cox, total_votes, pct_hogan, pct_cox)
hogan_v_cox
#Export that joined dataframe to a CSV file using `write_csv`
write_csv(hogan_v_cox, "data/hogan_v_cox.csv")
```

**Q2.** You're interested in seeing how the methods of voting have changed in Maryland since 2018. Using the `md_votes_by_type.csv` that has data from the top of the ticket (governor in 2018 and 2022, president in 2020), make a chart showing the percentage of votes cast in each category - early voting, election day and mail - over the three years. You'll need to make those percentages in R and then export the CSV using `write_csv`, then make the chart in Datawrapper. Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.

HINT: In Datawrapper's "Check and Describe" tab, you can click on a column to get the option of "hiding" it in any visualization.

What do you think the story here is? Put the answer below along with the URL of your chart.

**A2.** 
**https://datawrapper.dwcdn.net/FYC8k/1/ - I found it interesting that there was such a stark drop in total votes following the pandemic - which seemed to result in the percentage of early voters being lower in 2022 than 2020 and even 2018. With that observation I made the headline "Early Voting Down Nearly 11% Post Pandemic".**
```{r}
#Load the md_votes_by_type.csv file from the data folder
md_votes_by_type <- read_csv("data/md_votes_by_type.csv")
md_votes_by_type <- md_votes_by_type |> 
  clean_names()
#Add columns to generate the percentage of votes cast in each category
md_votes_by_type <- md_votes_by_type |> 
  mutate(pct_early = (early/total)*100, pct_elect = (election_day/total)*100, pct_mail = (mail/total)*100) |>
  group_by(year)
md_votes_by_type
#Export that dataframe to a CSV file using `write_csv`
write_csv(md_votes_by_type, "data/md_votes_by_type.csv")
```


**Q3.** You want to know more about the geography of 911 overdose calls in Montgomery County, Maryland, in 2022. To do that, you'll make a choropleth map in Datawrapper. You will need to:

* Read the file into R and create columns for the number of calls in each zip code and the percentage of all calls that number represents and then export that CSV to use in your map.
* Search for Maryland zip codes (use the 2020 data) as the base map.
* Upload the CSV file.
* Change the tooltip to show both the number and the percentage (you can put them side by side or put one under the other). Make sure to properly format both numbers to make them readable.
* Add labels to your map that locate prominent places in the county.
* Have a title, description and source that tells readers about the data. The title and description should highlight the most newsworthy or important aspect.
* Make any other changes to data that helps you make a more accurate map.

What do you think the story here is, and what would you want to know more about? Put the answer below along with the URL of your map.

**A3.** 
**https://datawrapper.dwcdn.net/SOrn4/1/ - I based this headline and description off the highest outlier in the data. I would definitely be interested in how median household income in this county compares to these call rates**
```{r}
#Load the md_votes_by_type.csv file from the data folder
montgomery_2022_overdoses <- read_csv("data/montgomery_2022_overdoses.csv")
montgomery_2022_overdoses <- montgomery_2022_overdoses |> 
  clean_names()
montgomery_2022_overdoses
#create columns for the number of calls in each zip code and the percentage of all calls that number represents
montgomery_overdoses <- montgomery_2022_overdoses |> 
  group_by(zip) |> 
  summarize(total_calls = n()) |>
  mutate(pct_calls = (total_calls/659)*100) |>
  arrange(desc(total_calls))
montgomery_overdoses
#export that CSV to use in your map
write_csv(montgomery_overdoses, "data/montgomery_overdoses.csv")
```