---
title: "Rreverse_engineering_project"
author: "Pablo Suarez & Mikayla Roberts"
date: "2023-10-15"
output: html_notebook
---
We started by loading our libraries and settings in the notebook
```{r}
#load libraries and settings
library(tidyverse)
library(janitor)
library(dplyr)
```
We then imported our data set into RStudio.

```{r}
#load data set
sports_political_donations <- read_csv("data/sports-political-donations.csv")

sports_political_donations
```
Then we cleaned the data set so that all the column names were uniform, the 'recipient' names were all upper case, and the 'amount' column was a number set rather than a character set.
```{r}
clean_sports_pol_donations <- sports_political_donations |>
  clean_names() |>
  mutate(recipient = str_to_upper(recipient))

clean_sports_pol_donations
```
Which team owners gave the most?
We need to single out donations from team owners by cycle and party donated to. We were able to almost completely replicate the data presented however, we were unable to trouble-shoot the 'election_cycle' column.
```{r}
clean_sports_pol_donations |>
  group_by(owner, team, league, party) |>
  summarize(total_amount = sum(amount)) |>
  arrange(desc(total_amount))
```
MLB owners have donated the most
To answer the question of "Which league owners have the most political donations?", we need to identify contributions from owners in each sport, by party, from 2016 to 2020. We were able to replicate the dataset however, there were a few total differences than that which is presented in the article.
```{r}
#group Republican donations by each league
mlb_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "MLB")) |>
  summarise(total = sum(amount))

nba_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "NBA")) |>
  summarise(total = sum(amount))

nhl_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "NHL")) |>
  summarise(total = sum(amount))

nfl_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "NFL")) |>
  summarise(total = sum(amount))

wnba_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "WNBA")) |>
  summarise(total = sum(amount))

nascar_republican <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Republican" & str_detect(league, "NASCAR")) |>
  summarise(total = sum(amount))

#group Democrat donations by each league
mlb_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "MLB")) |>
  summarise(total = sum(amount))

nba_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "NBA")) |>
  summarise(total = sum(amount))

nhl_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "NHL")) |>
  summarise(total = sum(amount))

nfl_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "NFL")) |>
  summarise(total = sum(amount))

wnba_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "WNBA")) |>
  summarise(total = sum(amount))

nascar_democrat <- clean_sports_pol_donations |>
  group_by(party, league) |>
  filter(party == "Democrat" & str_detect(league, "NASCAR")) |>
  summarise(total = sum(amount))

#combining each league's data into one dataset
bipartisan_donations <- data.frame (league  = c("MLB", "NBA", "NHL", "NFL", "WNBA", "NASACAR"),
    to_republicans = c(sum(mlb_republican$total), sum(nba_republican$total), sum(nhl_republican$total), sum(nfl_republican$total), sum(wnba_republican$total), sum(nascar_republican$total)),
    to_democrats = c(sum(mlb_democrat$total), sum(nba_democrat$total), sum(nhl_democrat$total), sum(nfl_democrat$total), sum(wnba_democrat$total), sum(nascar_democrat$total)))

bipartisan_donations <- bipartisan_donations |>
  mutate(total = to_republicans + to_democrats)

bipartisan_donations
```
"Significantly more money goes to Republicans: Specifically partisan contributions from owners and commissioners in the NFL, NBA, WNBA, NHL, MLB and NASCAR, per election cycle"
For this chart, we will need to group contributions by election year, and then create columns for each year that break down donations by party and then also show the total amount donated that year. This can be achieved by making separate dataframes for each party's donations and then merging the dataframes.

```{r}
#group Republican donation data
rep_df <- clean_sports_pol_donations |>
  group_by(election_year, party) |>
    summarise(total_amount=sum(amount)) |>
    filter(party=="Republican") |>
    arrange(desc(total_amount))

rep_df_clean = subset(rep_df, select = -c(party))

#group Democrat donation data
dem_df <- clean_sports_pol_donations |>
  group_by(election_year, party) |>
    summarise(total_amount=sum(amount)) |>
    filter(party=="Democrat") |>
    arrange(desc(total_amount))

dem_df_clean = subset(dem_df, select = -c(party))

#name Republican and Democrat columns
rep_df_clean <- rep_df_clean |>
  rename("to_republicans" = "total_amount")

dem_df_clean <- dem_df_clean |>
  rename("to_democrats" = "total_amount")

#combine Republican and Democrat data
year_party_donations <- merge(rep_df_clean, dem_df_clean, by = "election_year")

year_party_donations <- year_party_donations |>
  mutate(total = to_democrats + to_republicans) |>
  arrange(desc(election_year))

year_party_donations
```